---
title: "Biostat 203B Homework 1"
author: Rayo Suseno
subtitle: Due Jan 21 @ 11:59PM
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Display machine information for reproducibility:
```{r, eval=T}
sessionInfo()
```

## Q1. Git/GitHub

**No handwritten homework reports are accepted for this course.** We work with Git and GitHub. Efficient and abundant use of Git, e.g., frequent and well-documented commits, is an important criterion for grading your homework.

1. Apply for the [Student Developer Pack](https://education.github.com/pack) at GitHub using your UCLA email. You'll get GitHub Pro account for free (unlimited public and private repositories).

2. Create a **private** repository `biostat-203b-2022-winter` and add `Hua-Zhou` and `maschepps` as your collaborators with write permission.

3. Top directories of the repository should be `hw1`, `hw2`, ... Maintain two branches `main` and `develop`. The `develop` branch will be your main playground, the place where you develop solution (code) to homework problems and write up report. The `main` branch will be your presentation area. Submit your homework files (R markdown file `Rmd`, `html` file converted from R markdown, all code and extra data sets to reproduce results) in `main` branch.

4. After each homework due date, teaching assistant and instructor will check out your main branch for grading. Tag each of your homework submissions with tag names `hw1`, `hw2`, ... Tagging time will be used as your submission time. That means if you tag your `hw1` submission after deadline, penalty points will be deducted for late submission.

5. After this course, you can make this repository public and use it to demonstrate your skill sets on job market.

## Q2. Data ethics training

This exercise (and later in this course) uses the [MIMIC-IV data](https://mimic-iv.mit.edu), a freely accessible critical care database developed by the MIT Lab for Computational Physiology. Follow the instructions at <https://mimic.mit.edu/docs/gettingstarted/> to (1) complete the CITI `Data or Specimens Only Research` course and (2) obtain the PhysioNet credential for using the MIMIC-IV data. Display the verification links to your completion report and completion certificate here. (Hint: The CITI training takes a couple hours and the PhysioNet credentialing takes a couple days; do not leave it to the last minute.)

**Solution:**

CITI Completion Report link: https://www.citiprogram.org/verify/?k2a099f0f-f293-4d18-855f-f2f487676bf2-46544187

CITI Completion Certificate link: https://www.citiprogram.org/verify/?w52dcb34e-26c2-4c4b-8088-a58a6cd8c732-46544187

## Q3. Linux Shell Commands

1. The `/mnt/mimiciv/1.0` folder on teaching server contains data sets from MIMIC-IV. Refer to the documentation <https://mimic.mit.edu/docs/iv/> for details of data files.  
    ```{bash}
    ls -l /mnt/mimiciv/1.0
    ```
Please, do **not** put these data files into Git; they are big. Do **not** copy them into your directory. Do **not** decompress the gz data files. These create unnecessary big files on storage and are not big data friendly practices. Just read from the data folder `/mnt/mimiciv/1.0` directly in following exercises. 

    Use Bash commands to answer following questions.

2. Display the contents in the folders `core`, `hosp`, `icu`. Why are these data files distributed as `.csv.gz` files instead of `.csv` (comma separated values) files? Read the page <https://mimic.mit.edu/docs/iv/> to understand what's in each folder.

**Solution:** Content of folder 'core' is
```{bash}
ls -l /mnt/mimiciv/1.0/core
```

Content of the 'hosp' folder is
```{bash}
ls -l /mnt/mimiciv/1.0/hosp
```

Content of the 'icu' folder is
```{bash}
ls -l /mnt/mimiciv/1.0/icu
```

The files are distributed as .csv.gz because their sizes are too big.
The .gz format indicates that these folders have been compressed/zipped.
Therefore, storing them zipped is a wiser option.

3. Briefly describe what bash commands `zcat`, `zless`, `zmore`, and `zgrep` do.

**Solution**

zcat: let you view a compressed .gz file

zless: let you browse through the .gz file flexibly

zmore: let you browse through the .gz file only moving forward

zgrep: let you search using a particular filter in a .gz file

4. What's the output of following bash script?
    ```{bash, eval=F}
    for datafile in /mnt/mimiciv/1.0/core/*.gz
      do
        ls -l $datafile
      done
    ```
Display the number of lines in each data file using a similar loop.

**Solution**

The bash script above will show all the files inside the 'core' directory that
ends with a .gz (in other words, all the compressed files inside 'core')

To display the number of lines in each data file, we can use:
```{bash}
for datafile in /mnt/mimiciv/1.0/core/*.gz
  do
    zcat $datafile | wc -l
  done
```

5. Display the first few lines of `admissions.csv.gz`. How many rows are in this data file? How many unique patients (identified by `subject_id`) are in this data file? (Hint: combine Linux commands `zcat`, `head`/`tail`, `awk`, `sort`, `uniq`, `wc`, and so on.)

**Solution**

Display the first few lines of admission:
```{bash}
zcat /mnt/mimiciv/1.0/core/admissions.csv.gz | head
```

Number of rows in this data file:
```{bash}
zcat /mnt/mimiciv/1.0/core/admissions.csv.gz | wc -l
```

Number of unique patients:
```{bash}
zcat /mnt/mimiciv/1.0/core/admissions.csv.gz | uniq -w 8 | wc -l
```

6. What are the possible values taken by each of the variable `admission_type`, `admission_location`, `insurance`, and `ethnicity`? Also report the count for each unique value of these variables. (Hint: combine Linux commands `zcat`, `head`/`tail`, `awk`, `uniq -c`, `wc`, and so on.)

**Solution**
```{bash}
#zcat /mnt/mimiciv/1.0/core/admissions.csv.gz | uniq -f 6 | head
```


## Q4. Who's popular in Price and Prejudice

1. You and your friend just have finished reading *Pride and Prejudice* by Jane Austen. Among the four main characters in the book, Elizabeth, Jane, Lydia, and Darcy, your friend thinks that Darcy was the most mentioned. You, however, are certain it was Elizabeth. Obtain the full text of the novel from <http://www.gutenberg.org/cache/epub/42671/pg42671.txt> and save to your local folder. 
    ```{bash, eval=FALSE}
    wget -nc http://www.gutenberg.org/cache/epub/42671/pg42671.txt
    ```
Explain what `wget -nc` does. Do **not** put this text file `pg42671.txt` in Git. Complete the following loop to tabulate the number of times each of the four characters is mentioned using Linux commands.
    ```{bash}
    wget -nc http://www.gutenberg.org/cache/epub/42671/pg42671.txt
    for char in Elizabeth Jane Lydia Darcy
    do
      echo $char:
      # some commands here
      grep $char pg42671.txt | wc -l
    done
    ```
**Solution**
The command wget -nc will not re-download the file from the link if file has 
already existed


2. What's the difference between the following two commands?
    ```{bash eval=FALSE}
    echo 'hello, world' > test1.txt
    ```
    and
    ```{bash eval=FALSE}
    echo 'hello, world' >> test2.txt
    ```

**Solution**

The '>' will write "hello, world" to test1.txt. If test1.txt already had
something written on it, it will be replaced with "hello, world".

The '>>' on the other hand will simply append "hello, world" to test2.txt.
Therefore, if there exist a previous content in test2.txt, it will not be
replaced.

3. Using your favorite text editor (e.g., `vi`), type the following and save the file as `middle.sh`:
    ```{bash eval=FALSE}
    #!/bin/sh
    # Select lines from the middle of a file.
    # Usage: bash middle.sh filename end_line num_lines
    head -n "$2" "$1" | tail -n "$3"
    ```
Using `chmod` make the file executable by the owner, and run 
    ```{bash eval=FALSE}
    ./middle.sh pg42671.txt 20 5
    ```
Explain the output. Explain the meaning of `"$1"`, `"$2"`, and `"$3"` in this shell script. Why do we need the first line of the shell script?

**Solution**
After creating the file, I changed the file permissions using
```{bash eval=FALSE}
chmod u+x middle.txt
```
The script first takes the first 20 lines of the file pg42671.txt, and then
pipes it to a command that only take the last 5 lines. So in conclusion, the
output is line 15-20 of the given text file.

The `"$1"` represents the first argument of the middle.sh bash script that we
devised, `"$2"` represents the second argument, and the `"$3"` represents the
third argument needed for the middle.sh script. In our particular example,
`"$1"` will be understood as "pg42671.txt" (string), `"$2"` is 20 (integer),
and `"$3"` is 5 (integer).


### Q5. More fun with Linux

Try following commands in Bash and interpret the results: `cal`, `cal 2021`, `cal 9 1752` (anything unusual?), `date`, `hostname`, `arch`, `uname -a`, `uptime`, `who am i`, `who`, `w`, `id`, `last | head`, `echo {con,pre}{sent,fer}{s,ed}`, `time sleep 5`, `history | tail`.

**Solution**

`cal` -> show current calendar and indicate current date

`cal 2021` -> show the calendar for the year 2021

`cal 9 1752` -> show the calendar for the 9th month (September) of the year 1752
.For some reason, the command "cal 9 1752" does not show the date from the 3rd
until the 13th of September. 

`date` ->  gives us the current time and date

`hostname` -> gives us the name of the server hosting this environment

`arch` -> gives us the processor-bit information 

`uname -a` -> gives us complete system information which includes OS kernel
name, hostname, kernel release, details like time the last kernel was compiled,
machine architecture, processor architecture, OS' architecture, OS

`uptime` -> tells us how long has the system been running as well as how many
users are running it now

`whoami` -> return our username in the server

`who` -> tells you which users are logged in to the server

`w` -> tells you information about when a user logged in, how long have they
been idle, as well as what they do/run

`id` -> tells you the different ids (uid, gid, groups, context) that you might
have

`last | head` -> display the latest users that logged in to the server

`echo {con,pre}{sent,fer}{s,ed}` -> print all the different possible words
that can be constructed with the given sub-words

`time sleep 5` -> pauses the execution of command for a given time. In this
particular example, it's going to pause for 5 seconds

`history | tail` -> takes in the history of our command lines and pipe it into
the tail command, which only display the last 10 command lines executed